\section{Introduction}

Many problems are naturally expressed in sorted first-order logic.
For example, to model problems in geometry, we may wish to divide 
the universe into objects that are points, planes or lines.
While the expressive power of sorted first-order logic is equivalent to that of
classical first-order logic, it is convenient and intuitive to the human user.
Unfortunately, none of the major existing theorem-provers and model finders
of today deal with sorts. We present in this paper a new tool, Monopoly,
which translates problems in sorted first-order logic to logically 
equivalent unsorted problems. The translated problem can then be passed on to
a theorem prover or model finder of our choice, which can then solve it for us.

\begin{example}[werewolves and villagers] As a running example, we
model a village that has been infested with werewolves.\footnote{Our
example is an oversimplification of the party game Werewolf, also
known as Mafia.} The werewolves only attack at night, and during the
day they are ordinary villagers, so no-one knows who they are. In
desperation, the villagers pick some unfortunate people to be burnt at
the stake. The village is saved if this action kills all the
werewolves.

We can model this situation with the aid of two sorts $\villager$ and
$\werewolf$. We have a predicate $\slain$ that characterises which
villagers were burnt alive and a function $\alterEgo$ that says who
each werewolf is during the day:
\label{ex:werewolf1}
\begin{align*}
\slain & \in \villager \to o \\ 
\alterEgo & \in \werewolf \to \villager
\intertext{To make the problem more interesting for Monopoly, we
only allow the villagers to kill at most one person, a \emph{victim}
of their choosing:}
\victim & \in \villager 
\end{align*}
We state that the $\alterEgo$ function must be injective and no-one other
than the victim may be killed:
\begin{align*}
\forall X,Y \in \werewolf \ & ( \alterEgo (X) = \alterEgo (Y) \Longrightarrow  X = Y) \\
\forall X \in \villager\ & (\slain(X) \Longrightarrow X = \victim) \\
\intertext{Finally, we assert that the villagers have won---all
  werewolves were slain:}
 % & \slain(\victim) \\
 \forall X \in \werewolf \ & (\alterEgo(X) = \victim)
\end{align*} 
\end{example}

We want to use a model finder on this example to see if it is possible
for the villagers to win the game. But model finders such as Paradox
only operate on unsorted first-order logic! Luckily, we can use
Monopoly to safely remove all the sorts from our formula and we get:
\begin{align*}
\forall X, Y \ & ( \werewolfp(X) \land \werewolfp(Y) \land \alterEgo(X) = \alterEgo(Y) \Longrightarrow X = Y) \\
\forall X & (\slain(X) \Longrightarrow X = \victim) \\
\forall X \ & (\werewolfp(X) \Longrightarrow \alterEgo(X) = \victim) \\
& \werewolfp(\skolemconstant)
\end{align*}
We can see that Monopoly has introduced a predicate $\werewolfp$ to
model the $\werewolf$ sort, but hasn't introduced a corresponding
predicate for villagers. Often we \emph{do} need to introduce extra
predicates but blah blah blah

Monopoly has introduced a predicate $\werewolfp$ to ensure soundness,
but has detected that no such predicate is needed for the villagers.

It is in general unsound to simply remove the sorts from a problem;
in the worst case, Monopoly needs to introduce a predicate for each
sort like it did with $\werewolfp$ above. 

often, we need to introduce a predicate that characterises each of the
sorts. However, this introduces extra overhead 

blah blah blah

The reason why we need a predicate for werewolves but not for
villagers has to do with \emph{monotonicity}. If we think of an
interpretation of our in a domain of size $n$, we are free to have
fewer than $n$ werewolves in our interpretation, since only those
domain elements $x$ for which $\werewolfp(x)$ is true are
werewolves. However, we have no choice but to have $n$ villagers,
because each domain element is unconditionally a villager. So by not
including a $\villagerp$ predicate, we lose some control over

What we can do though is to translate sorted logic to unsorted logic. 
(this is possible since sorted logic and unsorted logic are equally expressive) 
We can then use an existing model finder (or theorem prover) for unsorted logic.

How to translate?

To simply remove the sorts is unsound, as we shall see in section ...
Introduce sorting predicates - inefficient.

Use monotonicity to infer when it is necessary to introduce sorting predicates
and when it is safe to remove the sorts. (Mention Jasmins paper)

- were going to show you lots of cool stuff in the next sections. blahblah.



%TODO cite Enderton 1972 - translation sorted - unsorted by predicates.










